Designing a high-performance boundary element library with OpenCL and Numba

### 2021-04-06 bempp_overview.tex
make ise, isation, etc consistent

Once the user defines an operator with its associated discretisation spaces and calls the \textbf{weak\_form} method to discretize it the code calls a regular integrator to assemble all the interactions between non-adjacent elements and a singular integrator to compute the interactions between adjacent triangles (only necessary if trial and test space are defined on the same grid). The result is stored internally as a dense Numpy array. The corresponding discretisation routines are proxies that forward to computational routines in either Numba or OpenCL, depending on the user preferences. For OpenCL assembly the code checks additional parameters, such as the default vector length for SIMD operators (e.g. 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), and whether the discretisation should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretisation routines and executed. For simple piecewise constant function spaces or other spaces, where the support of each basis function is localized to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

>>>>

Once the user defines an operator with its associated discretisation spaces and calls the \textbf{weak\_form} method to discretise it the code calls a regular integrator to assemble all the interactions between non-adjacent elements and a singular integrator to compute the interactions between adjacent triangles (only necessary if trial and test space are defined on the same grid). The result is stored internally as a dense Numpy array. The corresponding discretisation routines are proxies that forward to computational routines in either Numba or OpenCL, depending on the user preferences. For OpenCL assembly the code checks additional parameters, such as the default vector length for SIMD operators (e.g. 4 for double precision and 8 for single precision in Intel AVX2, or 1 if a GPU is used), and whether the discretisation should proceed in single or double precision. The OpenCL kernel is then compiled for the underlying compute device using PyOpenCL and executed. If the computational backend is Numba, the call is forwarded to the corresponding Numba discretisation routines and executed. For simple piecewise constant function spaces or other spaces, where the support of each basis function is localised to a single triangle, only one call to the computational routines is necessary. If the support of basis functions is larger than a single triangle different threads may need to sum into the same global degree of freedom. We discuss this in more detail in \cref{sec:coloring}.

### 2021-04-06 opencl.tex
make ise, isation, etc consistent

OpenCL (\url{https://www.khronos.org/opencl/}) is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core OpenCL executes compute kernels that can be written in C99, or more recently, in C++. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses. OpenCL splits the computational tasks into work-items, each of which represents a single execution of a compute kernel. Work items are grouped together into work-groups, which share local memory. Barrier synchronization is only allowed within a work-group. All work-items are uniquely indexed by a one, two, or three dimensional index space, called NDRange. Kernels are launched onto a compute device (e.g. CPU or GPU) from a host device. OpenCL allows kernels to be loaded as strings and compiled on the fly for a given device, making it well suited for launching from high-productivity languages. To launch an OpenCL kernel the following steps need to be performed.
\begin{itemize}
	\item Initialize buffers on the compute device.
	\item Compute relevant data from host to the compute device.
	\item Load a kernel string and just-in-time compile it for the device.
	\item Execute the kernel for the given buffers and chosen index space.
	\item Copy results back to the host.
\end{itemize}
An excellent feature of OpenCL is its very good support for vectorized operations. OpenCL provides vector data types, such as \textbf{double4}, that can hold four double values in a SIMD register and defines a number of standard operations for these vector types. This makes it easy to explicitly target modern SIMD execution in a portable way while avoiding difficult compiler intrinsics and keeping kernel code readable.

Python has excellent OpenCL support through the PyOpenCL library by Andreas Kloeckner, which automates much of the initialization of the OpenCL environment and makes it easy to create buffers and launch OpenCL kernels from Python.

@@

\subsection{SIMD optimized kernels}

The above description was valid for kernels that do not take advantage of CPU SIMD operations are kernels that run directly on a GPU device. For SIMD optimization on CPUs we proceed slightly differently. The corresponding kernel in principle works as described above, but batches together the interaction on one test triangle with X trial triangles, where X is either 4, 8, or 16, depending on the number of available SIMD lanes. This strategy allows us to optimize almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorized kernel.

Each kernel definition is stored in two variants, one with the ending \textit{\_novec.cl} and another one with the ending \textit{\_vec.cl}. The vectorized variant is configured via preprocessor directives for the desired number of vector lanes. Having to develop two OpenCL kernel codes for each operator is a certain amount of overhead. If we write a new kernel, we first implement the non-vectorized version. Then, with the help of preprocessor directives and a number of helper functions that do the actual implementation of operations depending on whether the kernel is vectorized or not, it is usually a matter of an hour or two to convert the non-vectorized kernel into a vectorized version.

Alternatively, some CPU OpenCL runtime environments have the option to try and auto-vectorize kernels by batching together work items on SIMD lanes, similarly to what we do manually. In our experience this works well for very simple kernels but often fails for more complex OpenCL kernels. This is why we decided to implement this strategy manually.

A completely different SIMD strategy is based on batching together quadrature evaluations within a single test/triangle pair. There are two disadvantages to this approach. First, it only works well if the number of quadrature points is a multiple of the available SIMD lanes. Second, other operations such as the geometry calculations for each element then can not be SIMD optimized as these are only performed once per test/triangle pair.

>>>>

OpenCL (\url{https://www.khronos.org/opencl/}) is a heterogeneous compute standard for CPUs, GPUs, FPGAs, and other types of devices that provide conformant drivers. At its core OpenCL executes compute kernels that can be written in C99, or more recently, in C++. The current version of OpenCL is 3.0, though the most widely implemented standard is OpenCL 1.2, which Bempp-cl uses. OpenCL splits the computational tasks into work-items, each of which represents a single execution of a compute kernel. Work items are grouped together into work-groups, which share local memory. Barrier synchronisation is only allowed within a work-group. All work-items are uniquely indexed by a one, two, or three dimensional index space, called NDRange. Kernels are launched onto a compute device (e.g. CPU or GPU) from a host device. OpenCL allows kernels to be loaded as strings and compiled on the fly for a given device, making it well suited for launching from high-productivity languages. To launch an OpenCL kernel the following steps need to be performed.
\begin{itemize}
	\item Initialise buffers on the compute device.
	\item Compute relevant data from host to the compute device.
	\item Load a kernel string and just-in-time compile it for the device.
	\item Execute the kernel for the given buffers and chosen index space.
	\item Copy results back to the host.
\end{itemize}
An excellent feature of OpenCL is its very good support for vectorised operations. OpenCL provides vector data types, such as \textbf{double4}, that can hold four double values in a SIMD register and defines a number of standard operations for these vector types. This makes it easy to explicitly target modern SIMD execution in a portable way while avoiding difficult compiler intrinsics and keeping kernel code readable.

Python has excellent OpenCL support through the PyOpenCL library by Andreas Kloeckner, which automates much of the initialisation of the OpenCL environment and makes it easy to create buffers and launch OpenCL kernels from Python.

@@

\subsection{SIMD optimised kernels}

The above description was valid for kernels that do not take advantage of CPU SIMD operations are kernels that run directly on a GPU device. For SIMD optimisation on CPUs we proceed slightly differently. The corresponding kernel in principle works as described above, but batches together the interaction on one test triangle with X trial triangles, where X is either 4, 8, or 16, depending on the number of available SIMD lanes. This strategy allows us to optimise almost all floating point operations within a kernel run for SIMD operation. If the number of trial elements is not divisible by 4, 8, or 16, then the few remaining trial elements are assembled with the standard-non vectorised kernel.

Each kernel definition is stored in two variants, one with the ending \textit{\_novec.cl} and another one with the ending \textit{\_vec.cl}. The vectorised variant is configured via preprocessor directives for the desired number of vector lanes. Having to develop two OpenCL kernel codes for each operator is a certain amount of overhead. If we write a new kernel, we first implement the non-vectorised version. Then, with the help of preprocessor directives and a number of helper functions that do the actual implementation of operations depending on whether the kernel is vectorised or not, it is usually a matter of an hour or two to convert the non-vectorised kernel into a vectorised version.

Alternatively, some CPU OpenCL runtime environments have the option to try and auto-vectorise kernels by batching together work items on SIMD lanes, similarly to what we do manually. In our experience this works well for very simple kernels but often fails for more complex OpenCL kernels. This is why we decided to implement this strategy manually.

A completely different SIMD strategy is based on batching together quadrature evaluations within a single test/triangle pair. There are two disadvantages to this approach. First, it only works well if the number of quadrature points is a multiple of the available SIMD lanes. Second, other operations such as the geometry calculations for each element then can not be SIMD optimised as these are only performed once per test/triangle pair.
